---
title: "Prediction of exercise quality using models trained with accelerometers data "
author: "Joe"
date: "30th Janurary 2015"
output: html_document
---
## Executive Summary
This report proposes a model to predict the quality of exercise performed by individuals. Different models are trained using data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. Best model is chosen based on the accuracy. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways.  More information is available at here : http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).

The dataset has been preprocessed and cleaned of irrevelant/not available columns. Highly correlated predictors have been eliminated. Predictors have been reduced from 160 variables to 34 variables after preprocessing. 

As the prediction is a multi classification problem, the models chosen were as random forest and boosting which perform well on multi classification. The data was split into training and test set. Models were trained on training data with cross validation and finally the best model was chosen on the prediction on the testing set.

The best model was randomforest with an accuracy of 0.9998 and Kappa of 0.9997 even though boosting also performed very well. 
Using this model, we predicted the given test data for quality of the exercises.

## Data Analysis
```{r, cache=TRUE, echo=FALSE}
library(caret)
library(plyr)
library(dplyr)
set.seed(32343)
data <- read.csv("pml-training.csv")
testing <- read.csv("pml-testing.csv")
data_row <- nrow(data)
testing_row <- nrow(testing)
tr <- names(data)
te <- names(testing)
c(tr[which(tr != te)], te[which(tr != te)])
```

## Preprocessing
```{r, cache=TRUE, echo=FALSE}
# Removed duplicate date and irrevelant index column X
sum(format(as.POSIXct(data$raw_timestamp_part_1, origin="1970-01-01", tz="UTC"),"%d/%m/%Y %H:%M") != data$cvtd_timestamp)
data <- data[,c(-1,-5)]

# Remove columns that have no or near zero variability
nzv <- nearZeroVar(data)
data <- data[,-nzv]

# Checking for missing data
missing <- mean(is.na(data))
data_na_count <- sapply(data, function(y) sum(is.na(y) | y == ""))
count(data_na_count)
data <- data[,names(subset(data_na_count,data_na_count==0))]
te_complete <- testing[,names(testing) %in% names(data)]

# Convert character columns to numeric
data$user_name <- factor(data$user_name)
user_levels <- levels(data$user_name)
data$user_name <- as.numeric(data$user_name)
testing$user_name <- as.numeric(factor(testing$user_name, levels=user_levels))
data$classe <- as.numeric(factor(data$classe))

# correlation 
data_cor <- cor(data)
summary(data_cor[upper.tri(data_cor)])
highlyCor <- findCorrelation(data_cor, cutoff = .75)
data_complete <- data[,-highlyCor]
data_complete$classe <- factor(data_complete$classe)
```

## Model Selection
```{r, cache=TRUE, echo=FALSE}
# Split data into training and testing set. 
train_index <- createDataPartition(data_complete$classe,p=0.75, list=FALSE)
training <- data_complete[train_index,]
validation <- data_complete[-train_index,]


# Apply Models
library(parallel)
library(doParallel)
cluster <- makeCluster(detectCores() - 2) # convention to leave 1 cor e for OS
registerDoParallel(cluster)
fitControl <- trainControl(method = "cv", number = 10, allowParallel = TRUE)
model = readRDS("rfmodel.rds")
if (is.null(model)) {
        tm <- proc.time()
        model <- train(classe~., data=training, method="rf", trControl = fitControl)
        rftime <- proc.time() - tm
}

pred <- predict(model, newdata=validation)
confusionMatrix(pred, validation$classe)

gbmModel = readRDS("gbmModel.rds")
if (is.null(gbmModel)) {
        tm <- proc.time()        
        gbmModel <- train(classe~., data=training, method="gbm", trControl = fitControl)
        gbmtime <- proc.time()-tm
}
predGbm <- predict(gbmModel, newdata = validation)
confusionMatrix(predGbm, validation$classe)

stopCluster(cluster)
```
